Scaffolder - Software for Manual Genome Scaffolding.

### 2011-09-12 src/article.tex
Make list of abbreviations into paragraph
Add approval statement
Merge in all track changes

\title{Scaffolder - Software for \change{Reproducible}{Manual} Genome
Scaffolding.}
%{{{1

@@

\address{
  \iid(1) Biology Department, The University of Akron, Akron, %
  OH 44325-3908, USA \\
  \iid(2) Department of Biological Sciences, Northern Kentucky %
  University, Nunn Drive, Highland Heights, KY 41076, USA }%

@@

  \paragraph*{Background:} \add{The} assembly of \add{next-generation}
  short-read sequencing data can result in a fragmented non-contiguous
  \change{series}{set} of genomic sequences. Therefore a common step in
  a genome project is to join neighbouring sequence regions together and fill
  gaps \remove{in the assembly using additional sequences}. This scaffolding
  step\remove{, however,} is non-trivial and requires manually editing large
  blocks of nucleotide sequence. Joining these sequences together also hides
  the source of each region in the final genome sequence. Taken together,
  these considerations may make reproducing or editing an existing genome
  \change{build}{scaffold} difficult.

  \paragraph*{Methods:} The software outlined here, ``Scaffolder,'' is
  implemented in the Ruby programming language and can be installed via the
  RubyGems software management system. Genome scaffolds are defined using YAML
  - a data format\remove{,} which is both human and machine-readable. Command
  line binaries and extensive documentation are available.

  \paragraph*{Results:} This software allows a genome \change{scaffold}{build}
  to be defined in terms of the constituent sequences using a relatively simple
  syntax\remove{to define the scaffold}. This syntax further allows unknown
  regions to be \change{defined}{specified} and \remove{add} additional
  sequence to \add{be used} fill \add{known} gaps in the scaffold. Defining
  the genome construction in a file makes the scaffolding process reproducible
  and easier to edit compared with \add{large} FASTA nucleotide sequences.

  \paragraph*{Conclusions:} Scaffolder is easy-to-use genome scaffolding
  software\change{. This tool}{ which} promotes reproducibility and continuous
  development in a genome project. Scaffolder can be found at \scaffolder.

@@

Assembly software takes the nucleotide reads produced by sequencing hardware
and, in the ideal case, outputs a single complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle:
each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
\remove{within the genome, }or biased and incomplete sequencing data, may
prevent the genome being assembled into a continuous sequence. This may be due
to \change{not enough}{insufficient}, or multiple different overlaps between
reads and is analogous to missing pieces in the jigsaw or pieces that fit to
multiple other pieces. \pb

The advent of high\add{-}throughput sequencing methods has led to a renewed
interest in algorithms to solve the problem of genome assembly
\cite{pop2008,pop2009}. \remove{Nonetheless, due to }The complexity of merging
large numbers of overlapping reads \add{can lead to} genome assembly software
\remove{may} being unable to produce a complete sequence. Instead, the
algorithm may generate several large assembled regions of sequence (`contigs')
composed of the many individual reads\remove{ that represent assembly into the
longest possible sequences, given the data. Nonetheless, }These contigs
represent a fragmented picture of the genome and \add{therefore} require
additional work to join together into a complete sequence. \pb

The process of \change{ameliorating a draft sequence into a finished continuous
sequence}{finishing a genome sequence} can be expensive in terms of time and
laboratory effort. \change{although}{In some cases} the genomic data present in
a set of generated contigs may be sufficient for many research questions
\cite{branscomb2002}. Nevertheless, a continuous high-quality `finished' genome
sequence does provide a greater depth of information, such as complete
resolution of repeat regions and precise estimates of distances between genomic
elements \cite{parkhill2002,fraser2002}. The process of joining these contigs
together to form a continuous genome sequence is called the `scaffolding' or
`finishing' stage and is the focus of the software described in this article.
\pb

@@

The sequencing process generates reads from either strand of the DNA helix and
the resulting contigs constructed from these reads may represent either the
5'~$\rightarrow$~3' or 3'~$\rightarrow$~5' DNA strands with respect to the
origin of replication. Orientating all contigs to point in the same direction
requires reverse complementing sequences where necessary. In the case of
\change{microbial}{archaeal and bacterial} genomes this orientation will be to
the 5'~$\rightarrow$~3' direction following the direction of genome
replication.

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of any
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence. \change{which}{This sequence} is useful for representing
both the known and to-be-resolved areas in the genome sequence.

@@

During the scaffolding process, closing and filling gaps between contigs
completes and improves the genome scaffold. Closing gaps may require returning
to the laboratory to perform additional sequencing or using computational
methods of estimating the unknown sequence. This additional sequence is
\remove{then }used to replace the gap between two contigs, joining them into
a single sequence. Once all contigs have been joined and gaps in a scaffold
closed, the genome may be considered finished.

@@

The process of \change{resolving a genome sequence and building a}{finishing
a genome} scaffold \remove{may} uses wet laboratory methods, \emph{in silico}
methods, or a combination of both. An example of a computational method might
use the \remove{existing} paired-read data from the sequencing stage.
\remove{where reads were produced in pairs at a known distance apart in the
genome.} The occurrence of paired reads in separate contigs can be used to
\change{probabilistically estimate}{estimate probabilistically} the order and
distance between these contigs. Alternatively, laboratory methods may use PCR
to amplify the unknown DNA in a gap region then use \remove{traditional}
Sanger sequencing to determine the sequence of this gap. Computational methods
are \remove{more} preferable as they are less costly in laboratory time and
materials compared to manual gap resolution, and use the available data that
exists following sequencing \cite{nagarajan2010}. Finally when the scaffold
cannot be completely resolved, \emph{in silico} software packages exist to
suggest the likely primers necessary for PCR amplifying the sequence in gap
regions \cite{gordon2001}. \pb

Examples of \emph{in silico} methods include \remove{using synteny to}
comparing the assembled contigs to a complete reference genome sequence
\change{by}{to} search\remove{ing} for areas of sequence similarity between the
two. Any areas of corresponding sequence with the reference genome can be used
to infer contig placement and build the contigs into a scaffold
\cite{richter2007,zhao2008,assefa2009}. \add{Genomic} recombination
\remove{within each of the two compared genomes} can however reduce the
\change{efficiency}{efficacy} of this. \pb

Repeat regions may also be responsible for multiple gaps when building
a genome sequence; tandemly repeated nucleotide regions in the genome produce
multiple reads with similar sequence. As many assembly algorithms rely on
sequence overlaps between reads to build a contig, the similarity between
repeat-region reads can result in the assembly collapsing into an artificially
short sequence or being ignored by more conservative assembly algorithms. Such
regions can be resolved by using algorithms \add{that} specifically reassemble
the collapsed repeat region \remove{correctly} \cite{mulyukov2002,koren2010}.
\change{while a}{A} related approach uses unassembled sequence reads matching
the regions around a scaffold gap to construct a uniquely overlapping set of
reads across \change{it}{the gap}. \cite{tsai2010}. \pb

@@

These described \emph{in silico} methods provide a wide array of approaches for
merging contigs into a larger, continuous scaffold sequence.
\remove{Nevertheless} The scaffolding process may still require manually
inserting additional sequences or further joining contigs using PCR-derived
sequence. Moving and editing large blocks of nucleotide text by hand however
\remove{possibly} introduces human error and precludes any reproducibility of
the steps in this process. \pb

The software outlined here, ``Scaffolder,'' aims to address these problems of
reproducibility by creating a file syntax and software framework for editing
a genome scaffold. Scaffolder uses a specific file format to define how
contigs are joined, additional sequences are inserted, and for the
specification of unknown regions. This syntax allows a scaffold to be updated
by simply editing the scaffold file. As such, Scaffolder facilitates
a reproducible \change{scaffolding}{finishing} process and provides a concise
overview of how the final genomic scaffold was constructed. \pb

@@

Scaffolder is written in the Ruby programming language \change{using version
1.8.7}{and tested against versions 1.8.7 and 1.9.2} \cite{ruby-lang}. The
Scaffolder package is split into two libraries. The first called ``scaffolder''
which provides the core Scaffolder application programming interface (API). The
second library ``scaffolder-tools'' provides the Scaffolder command line
interface (CLI). \pb

@@

The input data for Scaffolder is nucleotide sequence in FASTA format file.
These nucleotide sequences can be of any length and for example may be
individual \change{sequences}{reads}, assembled contigs or contigs which have
been joined to into larger scaffolds. The case in which Scaffolder may be most
useful is using the contigs and scaffolded contigs produced from the
previously assembled set of sequencing reads, combined with additional gap
filling sequences produced by PCR or \emph{in silico} methods as outlined in
the Background. \pb

\remove{The scaffolder algorithm reads through the scaffolder file parsing.
Each entry is then converted to corresponding nucleotide sequence by fetching
from the FASTA file and then performing any designated sequence trimming or
reverse complementing. Each of the sequence entries are then joined into
a continuous single super-sequence and returned to the console output.}

@@

The Scaffolder software facilitates reproducibly joining nucleotide sequences
together into a single contiguous scaffolded super-sequence. Plain-text
scaffold files written in \remove{the} YAML \remove{syntax} specify how these
sequences should be joined\change{. and the}{The} Scaffolder software is used
to generate the scaffold sequence from these instructions. In addition to
specifying which contigs are required, the scaffold file allows the contigs to
be edited into smaller sub-sequences or reverse complemented if necessary.
\add{Each scaffold file represents one scaffolded nucleotide sequence and
separate scaffolds should be defined in separate files.} \pb

The process of genome finishing may involve producing additional
oligonucleotide sequences to fill unknown regions in \change{the}{a} scaffold.
\remove{and} The Scaffolder \change{allows}{format provides functionality to
use} these additional insert sequences \remove{be used} to fill \remove{these}
gaps. These inserts can also be treated in the same manner as larger contig
sequences: trimmed and/or reverse complemented to match the corresponding gap
region size and orientation. \pb

The distances between contigs may be estimated from paired-read data or from
mapping the contigs to a reference genome. These inter-contig gap regions are
useful to join separate sequences together by \change{an}{the} estimated
\change{size and }{distance.} The scaffold file allows for the specification
of such unresolved regions by inserting regions of `N' nucleotides into the
scaffold. The use of these regions in the scaffold indicates the unresolved
regions in the build and their approximate size. \pb

\change{The scaffold file specifies how sequence regions are joined together;
however, the sequences themselves are no  stored in the scaffold file and are
instead}{The nucleotide sequences used in the scaffold are} maintained as
a separate FASTA file. \remove{Nucleotide sequences in a FASTA-format file are
the standard output for a genome assembler, and these may be contigs, sets of
contigs scaffolded into larger sequences, or both. }The nucleotide sequences
\remove{stored in the FASTA file} are referenced in the scaffold using the
first word from \remove{in} the FASTA header of the corresponding sequence.
Maintaining the nucleotide sequences in a separate file preserves the unedited
sequence and decouples the data from the specification of how it should be
used to produce the genome sequence. \pb

@@

The first line of the scaffold file in Figure 1 begins with three dashes to
indicate the start of a YAML-formatted document. The first entry (highlighted
in green) begins with a dash character `-' to denote an entry in the YAML
list. This is a requirement of the YAML format: \remove{that} each entry
begins with a dash line. The next line is indented by two spaces where
whitespace is \change{required}{used} to group similar attributes together.
The ``sequence'' tag indicates that this entry corresponds to a sequence in
the FASTA file and the following line indicates the name of this sequence
using the ``source'' tag. The first word of the FASTA header is used to
identify which sequence is selected from the file. Together these three lines
describe the first entry in the scaffold as a simple sequence using the
sequence named `sequence1'. On the right hand side of Figure 1 this produces
the first region in the scaffold, also shown in green. \pb

@@

The last entry in the scaffold, highlighted in blue, adds the sequence named
`sequence2' to the scaffold. This entry demonstrates how this sequence may be
manipulated prior to addition to the scaffold. The `start' and `stop' tags
trim the sequence to these coordinates inclusively.\remove{while} The
``reverse'' tag \add{also} instructs Scaffolder to \remove{also} reverse
complement the sequence. In the putative scaffold shown in Figure 1 this
completes the \remove{scaffold } sequence. \pb

The final entry in the scaffold \change{also includes}{uses} the ``inserts''
tag to add additional regions of sequence. These inserts are also added as
a YAML list, with each insert beginning with a dash. The first insert, shown
in purple, uses similar attributes to that of a sequence entry; the reverse,
start and stop tags are used to trim and reverse complement the insert.
Similarly the `source' tag identifies the corresponding FASTA sequence as
`insert1'. The ``open'' and ``close'' tags are specific to inserts and
determine where the insert is added in the enclosing sequence. The region of
the sequence inside these coordinates is inclusively replaced by the specified
insert sequence. This is visualised in the putative scaffold in Figure 1 by
the black lines bisecting the blue sequence. \pb

@@

Scaffolder provides a standardised set of Ruby classes and methods (API) for
interacting with the scaffold. This allows Scaffolder to be integrated into
existing genomics workflows or used with Ruby build tools such as Rake. In
addition Scaffolder provides a command line interface (CLI) to validate the
scaffold file and build the draft super sequence. The Scaffolder CLI behaves
as a standard Unix tool and returns appropriate exit codes and manual pages.
The use of both these Scaffolder interfaces is outlined in detail on the
Scaffolder website (\scaffolder). \add{This website provides a ``getting
started'' guide as an introduction to using Scaffolder to build a genome
scaffold.} \pb

@@

Scaffolding an incomplete genome assembly requires joining contigs and
additional gap-filling sequence\add{s} using a combination of computational and
laboratory methods. The process of manually editing a scaffold is inherently
hard to reproduce and introduces \change{the possibility of irreducible of
techniques}{irreproducible edits} and/or human error. In respect to
\change{these drawbacks}{this} the aims of the Scaffolder software are
twofold: 1) to provide software that is easy to install and simplifies the
task of genome finishing; and 2) to facilitate reproducibility in the
scaffolding and finishing stage of a genome project. \pb

\remove{Scaffolder was designed to be as simple to use and, assuming the Ruby
and RubyGems software are present, can be installed in a single command line
step. This should negate any possible barriers to entry which require manual
compilation of source code.} Scaffolder \remove{also} uses a minimal and
compact syntax to describe how \change{the}{a} genome scaffold sequence should
be generated. This syntax is simple to \change{construct}{write} and edit
whilst being succinct and readable. \remove{The goal of this syntax is to make
genome scaffolds easy to write in a common text editor.}

\add{AGP is a similar format for describing scaffolds. This format can be used
to describe contig order and N-filled gap regions in a scaffold. The
advantages of the AGP format is that each contig entry is defined on a single
line which allows searching the scaffold using UNIX line-based tools. The
Scaffolder format in comparison in contrast is written in the standardised
YAML format and therefore accessible to the many languages which provide
parsers to this format. The Scaffolder format is provided with a tool
explicitly to produce the fasta sequence of scaffold specified by in the file.
The scaffold format further provides functions for trimming and replacing
regions of sequence using inserts.} \pb

@@

\add{A use case for Scaffolder is a combination of computational and manual
editing of a genome scaffold. We have used Scaffolder in our own genome
projects to create an initial scaffold from compuationally parsing the output
of \emph{in silico} scaffolding tools into YAML. This scaffold was then
manually updated as the scaffold was finished with additional gap-filling
sequences generated in the laboratory. This is example of the scaffolder
format being both compuationally tractable while being simple to edit
manually.} \pb

The YAML text format \add{also} allows comparison of differences between
scaffold builds using standard Unix tools such as diff. This \add{therefore}
makes scaffold files \add{amenable} to storage in version control systems and
allows genome finishers to use methods similar to those in software
development. \pb

@@

  \begin{description}
    \item[Project name:] Scaffolder v0.4.\change{1}{4}, Scaffolder Tools
    v0.1.\change{2}{3}
    \item[Project home page:] \scaffolder
    \item[Operating system:] Platform Independent. Tested on Mac OS X and
    Ubuntu.
    \item[Programming language:] Ruby \add{1.8.7 or 1.9.2}
    \item[Other requirements:] \change{RubyGems}{RubyGems package management
    software and the following libraries: BioRuby 1.4.x, confligliere
    0.1.x, ronn 0.7.x. A full list of development dependencies can be found in
    the Gemfile in the base directory of each project.}
    \item[License:] MIT \item[Any restrictions to use by non-academics:] None
    \end{description}

@@

  \begin{description}
    \item[API:] Application programming interface
    \item[CLI:] Command line interface
    \item[PCR:] Polymerase chain reaction
    \item[YAML:] YAML ain't markup language\cite{yaml}
  \end{description}

@@

MDB developed and maintains the Scaffolder tool. MDB and HAB wrote the
manuscript.

>>>>

\title{Scaffolder - Software for Manual Genome
Scaffolding.}
%{{{1

@@

\address{
\iid(1) Biology Department, The University of Akron, Akron, %
OH 44325-3908, USA \\
\iid(2) Department of Biological Sciences, Northern Kentucky %
University, Nunn Drive, Highland Heights, KY 41076, USA}

@@

  \paragraph*{Background:} The assembly of next-generation short-read
  sequencing data can result in a fragmented non-contiguous set of genomic
  sequences. Therefore a common step in a genome project is to join
  neighbouring sequence regions together and fill gaps . This scaffolding step
  is non-trivial and requires manually editing large blocks of nucleotide
  sequence. Joining these sequences together also hides the source of each
  region in the final genome sequence. Taken together, these considerations
  may make reproducing or editing an existing genome scaffold difficult.

  \paragraph*{Methods:} The software outlined here, ``Scaffolder,'' is
  implemented in the Ruby programming language and can be installed via the
  RubyGems software management system. Genome scaffolds are defined using
  YAML - a data format which is both human and machine-readable. Command line
  binaries and extensive documentation are available.

  \paragraph*{Results:} This software allows a genome build to be defined in
  terms of the constituent sequences using a relatively simple syntax. This
  syntax further allows unknown regions to be specified and additional sequence
  to be used fill known gaps in the scaffold. Defining the genome construction
  in a file makes the scaffolding process reproducible and easier to edit
  compared with large FASTA nucleotide sequences.

  \paragraph*{Conclusions:} Scaffolder is easy-to-use genome scaffolding
  software which promotes reproducibility and continuous development in a
  genome project. Scaffolder can be found at \scaffolder.

@@

Assembly software takes the nucleotide reads produced by sequencing hardware
and, in the ideal case, outputs a single complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle:
each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
or biased and incomplete sequencing data, may prevent the genome being
assembled into a continuous sequence. This may be due to insufficient, or
multiple different overlaps between reads and is analogous to missing pieces
in the jigsaw or pieces that fit to multiple other pieces. \pb

The advent of high-throughput sequencing methods has led to a renewed interest
in algorithms to solve the problem of genome assembly \cite{pop2008,pop2009}.
The complexity of merging large numbers of overlapping reads can lead to
genome assembly software being unable to produce a complete sequence. Instead,
the algorithm may generate several large assembled regions of sequence
(`contigs') composed of the many individual readsThese contigs represent a
fragmented picture of the genome and therefore require additional work to join
together into a complete sequence. \pb

The process of finishing a genome sequence can be expensive in terms of time
and laboratory effort. In some cases the genomic data present in a set of
generated contigs may be sufficient for many research questions
\cite{branscomb2002}. Nevertheless, a continuous high-quality `finished'
genome sequence does provide a greater depth of information, such as complete
resolution of repeat regions and precise estimates of distances between
genomic elements \cite{parkhill2002,fraser2002}. The process of joining these
contigs together to form a continuous genome sequence is called the
`scaffolding' or `finishing' stage and is the focus of the software described
in this article. \pb

@@

The sequencing process generates reads from either strand of the DNA helix and
the resulting contigs constructed from these reads may represent either the
5'~$\rightarrow$~3' or 3'~$\rightarrow$~5' DNA strands with respect to the
origin of replication. Orientating all contigs to point in the same direction
requires reverse complementing sequences where necessary. In the case of
archaeal and bacterial genomes this orientation will be to the
5'~$\rightarrow$~3' direction following the direction of genome replication.

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of any
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence. This sequence is useful for representing both the known
and to-be-resolved areas in the genome sequence.

@@

During the scaffolding process, closing and filling gaps between contigs
completes and improves the genome scaffold. Closing gaps may require returning
to the laboratory to perform additional sequencing or using computational
methods of estimating the unknown sequence. This additional sequence is used
to replace the gap between two contigs, joining them into a single sequence.
Once all contigs have been joined and gaps in a scaffold closed, the genome
may be considered finished.

@@

The process of finishing a genome scaffold uses wet laboratory methods,
\emph{in silico} methods, or a combination of both. An example of a
computational method might use the paired-read data from the sequencing stage.
The occurrence of paired reads in separate contigs can be used to estimate
probabilistically the order and distance between these contigs. Alternatively,
laboratory methods may use PCR to amplify the unknown DNA in a gap region then
use Sanger sequencing to determine the sequence of this gap. Computational
methods are preferable as they are less costly in laboratory time and materials
compared to manual gap resolution, and use the available data that exists
following sequencing \cite{nagarajan2010}. Finally when the scaffold cannot be
completely resolved, \emph{in silico} software packages exist to suggest the
likely primers necessary for PCR amplifying the sequence in gap regions
\cite{gordon2001}. \pb

Examples of \emph{in silico} methods include comparing the assembled contigs
to a complete reference genome sequence to search for areas of sequence
similarity between the two. Any areas of corresponding sequence with the
reference genome can be used to infer contig placement and build the contigs
into a scaffold \cite{richter2007,zhao2008,assefa2009}. Genomic recombination
can however reduce the efficacy of this. \pb

Repeat regions may also be responsible for multiple gaps when building a
genome sequence; tandemly repeated nucleotide regions in the genome produce
multiple reads with similar sequence. As many assembly algorithms rely on
sequence overlaps between reads to build a contig, the similarity between
repeat-region reads can result in the assembly collapsing into an artificially
short sequence or being ignored by more conservative assembly algorithms. Such
regions can be resolved by using algorithms that specifically reassemble the
collapsed repeat region \cite{mulyukov2002,koren2010}. A related approach uses
unassembled sequence reads matching the regions around a scaffold gap to
construct a uniquely overlapping set of reads across the gap. \cite{tsai2010}.
\pb

@@

These described \emph{in silico} methods provide a wide array of approaches for
merging contigs into a larger, continuous scaffold sequence. The scaffolding
process may still require manually inserting additional sequences or further
joining contigs using PCR-derived sequence. Moving and editing large blocks of
nucleotide text by hand however introduces human error and precludes any
reproducibility of the steps in this process. \pb

The software outlined here, ``Scaffolder,'' aims to address these problems of
reproducibility by creating a file syntax and software framework for editing a
genome scaffold. Scaffolder uses a specific file format to define how contigs
are joined, additional sequences are inserted, and for the specification of
unknown regions. This syntax allows a scaffold to be updated by simply editing
the scaffold file. As such, Scaffolder facilitates a reproducible finishing
process and provides a concise overview of how the final genomic scaffold was
constructed. \pb

@@

Scaffolder is written in the Ruby programming language and tested against
versions 1.8.7 and 1.9.2 \cite{ruby-lang}. The Scaffolder package is split into
two libraries. The first called ``scaffolder'' which provides the core
Scaffolder application programming interface (API). The second library
``scaffolder-tools'' provides the Scaffolder command line interface (CLI). \pb

@@

The input data for Scaffolder is nucleotide sequence in FASTA format file.
These nucleotide sequences can be of any length and for example may be
individual reads, assembled contigs or contigs which have been joined to into
larger scaffolds. The case in which Scaffolder may be most useful is using the
contigs and scaffolded contigs produced from the previously assembled set of
sequencing reads, combined with additional gap filling sequences produced by
PCR or \emph{in silico} methods as outlined in the Background. \pb

@@

The Scaffolder software facilitates reproducibly joining nucleotide sequences
together into a single contiguous scaffolded super-sequence. Plain-text
scaffold files written in YAML specify how these sequences should be joinedThe
Scaffolder software is used to generate the scaffold sequence from these
instructions. In addition to specifying which contigs are required, the
scaffold file allows the contigs to be edited into smaller sub-sequences or
reverse complemented if necessary. Each scaffold file represents one scaffolded
nucleotide sequence and separate scaffolds should be defined in separate files.
\pb

The process of genome finishing may involve producing additional
oligonucleotide sequences to fill unknown regions in a scaffold. The Scaffolder
format provides functionality to use these additional insert sequences to fill
gaps. These inserts can also be treated in the same manner as larger contig
sequences: trimmed and/or reverse complemented to match the corresponding gap
region size and orientation. \pb

The distances between contigs may be estimated from paired-read data or from
mapping the contigs to a reference genome. These inter-contig gap regions are
useful to join separate sequences together by the estimated distance. The
scaffold file allows for the specification of such unresolved regions by
inserting regions of `N' nucleotides into the scaffold. The use of these
regions in the scaffold indicates the unresolved regions in the build and their
approximate size. \pb

The nucleotide sequences used in the scaffold are maintained as a separate
FASTA file. The nucleotide sequences are referenced in the scaffold using the
first word from the FASTA header of the corresponding sequence. Maintaining the
nucleotide sequences in a separate file preserves the unedited sequence and
decouples the data from the specification of how it should be used to produce
the genome sequence. \pb

@@

The first line of the scaffold file in Figure 1 begins with three dashes to
indicate the start of a YAML-formatted document. The first entry (highlighted
in green) begins with a dash character `-' to denote an entry in the YAML list.
This is a requirement of the YAML format: each entry begins with a dash line.
The next line is indented by two spaces where whitespace is used to group
similar attributes together. The ``sequence'' tag indicates that this entry
corresponds to a sequence in the FASTA file and the following line indicates
the name of this sequence using the ``source'' tag. The first word of the FASTA
header is used to identify which sequence is selected from the file. Together
these three lines describe the first entry in the scaffold as a simple sequence
using the sequence named `sequence1'. On the right hand side of Figure 1 this
produces the first region in the scaffold, also shown in green. \pb

@@

The last entry in the scaffold, highlighted in blue, adds the sequence named
`sequence2' to the scaffold. This entry demonstrates how this sequence may be
manipulated prior to addition to the scaffold. The `start' and `stop' tags trim
the sequence to these coordinates inclusively. The ``reverse'' tag also
instructs Scaffolder to reverse complement the sequence. In the putative
scaffold shown in Figure 1 this completes the sequence. \pb

The final entry in the scaffold uses the ``inserts'' tag to add additional
regions of sequence. These inserts are also added as a YAML list, with each
insert beginning with a dash. The first insert, shown in purple, uses similar
attributes to that of a sequence entry; the reverse, start and stop tags are
used to trim and reverse complement the insert. Similarly the `source' tag
identifies the corresponding FASTA sequence as `insert1'. The ``open'' and
``close'' tags are specific to inserts and determine where the insert is added
in the enclosing sequence. The region of the sequence inside these coordinates
is inclusively replaced by the specified insert sequence. This is visualised in
the putative scaffold in Figure 1 by the black lines bisecting the blue
sequence. \pb

@@

Scaffolder provides a standardised set of Ruby classes and methods (API) for
interacting with the scaffold. This allows Scaffolder to be integrated into
existing genomics workflows or used with Ruby build tools such as Rake. In
addition Scaffolder provides a command line interface (CLI) to validate the
scaffold file and build the draft super sequence. The Scaffolder CLI behaves as
a standard Unix tool and returns appropriate exit codes and manual pages. The
use of both these Scaffolder interfaces is outlined in detail on the Scaffolder
website (\scaffolder). This website provides a ``getting started'' guide as an
introduction to using Scaffolder to build a genome scaffold. \pb

@@

Scaffolding an incomplete genome assembly requires joining contigs and
additional gap-filling sequences using a combination of computational and
laboratory methods. The process of manually editing a scaffold is inherently
hard to reproduce and introduces irreproducible edits and/or human error. In
respect to this the aims of the Scaffolder software are twofold: 1) to provide
software that is easy to install and simplifies the task of genome finishing;
and 2) to facilitate reproducibility in the scaffolding and finishing stage of
a genome project. \pb

Scaffolder uses a minimal and compact syntax to describe how a genome scaffold
sequence should be generated. This syntax is simple to write and edit whilst
being succinct and readable.

AGP is a similar format for describing scaffolds. This format can be used to
describe contig order and N-filled gap regions in a scaffold. The advantages of
the AGP format is that each contig entry is defined on a single line which
allows searching the scaffold using UNIX line-based tools. The Scaffolder
format in comparison in contrast is written in the standardised YAML format and
therefore accessible to the many languages which provide parsers to this
format. The Scaffolder format is provided with a tool explicitly to produce the
fasta sequence of scaffold specified by in the file. The scaffold format
further provides functions for trimming and replacing regions of sequence using
inserts. \pb

@@

A use case for Scaffolder is a combination of computational and manual editing
of a genome scaffold. We have used Scaffolder in our own genome projects to
create an initial scaffold from compuationally parsing the output of \emph{in
silico} scaffolding tools into YAML. This scaffold was then manually updated as
the scaffold was finished with additional gap-filling sequences generated in
the laboratory. This is example of the scaffolder format being both
compuationally tractable while being simple to edit manually. \pb

The YAML text format also allows comparison of differences between scaffold
builds using standard Unix tools such as diff. This therefore makes scaffold
files amenable to storage in version control systems and allows genome
finishers to use methods similar to those in software development. \pb

@@

  \begin{description}
    \item[Project name:] Scaffolder v0.4.4, Scaffolder Tools
    v0.1.3
    \item[Project home page:] \scaffolder
    \item[Operating system:] Platform Independent. Tested on Mac OS X and
    Ubuntu.
    \item[Programming language:] Ruby 1.8.7 or 1.9.2
    \item[Other requirements:] RubyGems package management
    software and the following libraries: BioRuby 1.4.x, confligliere
    0.1.x, ronn 0.7.x. A full list of development dependencies can be found in
    the Gemfile in the base directory of each project.
    \item[License:] MIT \item[Any restrictions to use by non-academics:] None
    \end{description}

@@

API: Application programming interface, CLI: Command line interface, PCR:
Polymerase chain reaction, YAML: YAML ain't markup language\cite{yaml}.

@@

MDB developed and maintains the Scaffolder tool. MDB and HAB wrote the
manuscript. All authors have read and approved the manuscript.
